---
title: "Intervention Analysis of UKDriverDeaths Dataset"
author: "Tsu-Hao Fu"
output:
  pdf_document: default
date: "2024-4-7"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!require(forecast)) install.packages("forecast")
library(forecast)
if(!require(tseries)) install.packages("tseries")
library(tseries)
if(!require(TSA)) install.packages("TSA")
library(TSA)
```
## 1. Summary
We consider the monthly time series UKDriverDeaths containing the well-known data
from Harvey and Durbin (1986) on car drivers killed or seriously injured in the 
United Kingdom from January 1969  through December 1984. These are also known as
the “seatbelt data”, as they were used by Harvey and Durbin (1986) for evaluating
the effectiveness of compulsory wearing of seatbelts introduced on 1983-01-31. 
Intervention analysis format will be used in the proceeding sections.

```{r, warning=FALSE, echo = FALSE}
# Load the data
data(UKDriverDeaths)
# Plot the time series
plot(UKDriverDeaths, main = "Monthly Driver Deaths in Great Britain (1969-1984)"
     , ylab = "Number of Deaths")

```

The seatbelt data reveal clear trends and seasonality, necessitating the removal
of these elements to achieve stationarity for accurate forecasting of future 
driver fatalities. A log transformation is advisable to stabilize potential 
non-constant variance and improve the normality of the data. The red line on the
plot smooths out the data, elucidating underlying patterns by minimizing the 
impact of seasonal spikes. The green line marks the seat belt legislation, after
which a significant decrease in the data's level suggests the law's 
effectiveness in lowering fatalities and injuries. This observed downward trend 
underscores the need for differencing the log-transformed data to achieve 
stationarity, a critical step for subsequent forecasting.

```{r, echo=FALSE}
# Log Transformation
plot(log(UKDriverDeaths), main = "Log Monthly Driver Deaths in Great Britain 
     (1969-1984)", ylab = "Log Number of Deaths")
abline(v=c(1983 + (1-1)/12), col="green", lwd=1)
lines(filter(log(UKDriverDeaths), c(1/2, rep(1, 11), 1/2)/12),col = 2)
UKDriverDeaths = log(UKDriverDeaths)
```

Our final SARIMA model:
$\Delta \Delta_{12} Z_t = (1 - 0.6967B) (1 - 0.8985B^{12}) a_t - 0.2278 \cdot 
X_t$

where \( \Delta \Delta_{12} Z_t \) is the differenced time series at time t,
\( a_t \) is the white noise at time t and the intervention variable \(X_t\) 
represents the effect of the seat belt law implementation at time t.

This model is also famously known as the airline model due to its application to
a series of airline passengers in the classical text by Box and Jenkins (1970).


## 2. Analysis
## 2.1 EDA
By observing the data, we apply the additive decomposition since the seasonal 
effect remains constant over time and does not vary with the level of the time 
series.The trend component indicates there may be a long-term decrease in driver 
deaths over the period shown. The seasonal component captures regular 
fluctuations within a year, possibly related to seasonal factors affecting
driving conditions, such as ice or fog in winter months.

```{r, echo=FALSE}
# Decompose the time series into seasonal, trend, and random noises
plot(decompose(UKDriverDeaths, type = "additive"))
```
The t-test statistic is 8.5523 with a p-value much smaller than 0.05, indicating
a very significant difference in means before and after the law. The 95% 
confidence interval for the difference in means ranges from 302.9403 491.8350, 
suggesting the reduction in the number of deaths is statistically significant 
and not due to random chance.

```{r, echo=FALSE}
# Pre-Post Mean Comparison
pre_law <- window(UKDriverDeaths, end=c(1982, 12))
post_law <- window(UKDriverDeaths, start=c(1983, 2))
summary(pre_law)
summary(post_law)
t.test(exp(pre_law), exp(post_law))
```

## 2.2 Stationarity Testing
ADF Test: Indicates the series is stationary, suggesting no unit root is present.
This is a good sign for further analysis as many time series models require 
stationarity as a prerequisite.

KPSS Test for Level Stationarity: Suggests the series is not stationary when 
considering level stationarity. This could indicate the presence of a trend or 
changing variance over time.

KPSS Test for Trend Stationarity: Indicates that once a trend is considered, the
series does not appear to have a unit root, suggesting it is trend stationary.

Given these results, the UKDriverDeaths series seems to be trend stationary. 
This means that the underlying process generating the data could have a 
deterministic trend component, and once that trend is accounted for, the series 
behaves in a stationary manner. 

```{r, warning=FALSE, echo=FALSE}
# Augmented Dickey-Fuller (ADF) test
adf.test(pre_law)

# Kwiatkowski-Phillips-Schmidt-Shin (KPSS) for Level Stationarity
kpss.test(pre_law, null = 'Level')

# Kwiatkowski-Phillips-Schmidt-Shin (KPSS) for Trend Stationarity
kpss.test(pre_law, null = 'Trend')

```
After the first difference is taken, the tests suggest that the time series 
does not have a unit root, thereby indicating its stationarity. From the plot, 
the log-transformed seatbelt data also appears to be stationary after the first 
differencing.

```{r, echo=FALSE, warning=FALSE}
cat("Number of differences required to achieve stationarity:",
    ndiffs(pre_law,test = 'kpss'),"\n")
cat("Number of differences required to achieve seasonally stationarity:",
    nsdiffs(pre_law), "\n")

#  Tests for the first difference
adf.test(diff(pre_law))
kpss.test(diff(pre_law), null = 'Level')
kpss.test(diff(pre_law), null = 'Trend')

# Plot the differenced series
plot(diff(pre_law), main="First Differenced Series", 
     ylab="Differenced Deaths",
     col="blue")
```

## 2.3 Model Specification

The use of Autocorrelation Function (ACF) and Partial Autocorrelation Function 
(PACF) correlograms helped choose the SARIMA model parameters (p, q) and (P,Q) 
for the data. The observations within the 95% significance bounds generally 
indicate a whitenoise process. 

The first-differenced ACF plot reveals significant autocorrelations at regular 
intervals, specifically at lags 12, 24, and 36, suggesting a potential annual 
seasonality in the monthly data. Meanwhile, the first-differenced PACF plot shows
significant partial autocorrelations at the initial 13 lags, indicating a 
possible combination of autoregressive process and seasonality.

After double differencing, which includes a seasonal difference, the ACF and 
PACF plots still show  significant spikes at first 3 lags. These spikes could be
random or signify an important feature. Therefore, we assume that the optimal P 
and Q are around 3.

```{r, echo=FALSE}
# ACF and PACF plots
par(mfrow = c(2, 2))
Acf(diff(pre_law), main="ACF of First-Differenced Series",lag.max = 36)
Pacf(diff(pre_law), main="PACF of First-Differenced Series",lag.max = 36)

Acf(diff(diff(pre_law, differences=12)), main="ACF of Double-Differenced Series",
    lag.max = 36)
Pacf(diff(diff(pre_law, differences=12)), main="PACF of Double-Differenced Series",
     lag.max = 36)
```

## 2.4 Pre-Intervention Modelling

In order to select an appropriate SARIMA model that best fits the seatbelt 
data, a matrix of BIC values was used. BIC tends to favor simpler models than 
AIC, which can be an advantage for model interpretation and when the goal is to 
identify the true model. The BIC values suggested the $SARIMA (0,1,1)(0,1,1)_{12}$ 
model appeared to most adequately fit the seatbelt data.

```{r, warning=FALSE, echo=FALSE}
# Grid Search for the optimal SARIMA model
nd_pars <- expand.grid(ar = 0:2, diff = 1, ma = 0:2, sar = 0:2, sdiff = 1,
                       sma = 0:2)
nd_bic <- rep(Inf, nrow(nd_pars))

# Perform the grid search
for (i in seq_len(nrow(nd_pars))) {
  # Extract the parameters for this iteration
  pars <- nd_pars[i, ]

  # Fit the model within tryCatch to handle potential errors
  fit <- tryCatch({
    Arima(pre_law, order=c(pars$ar, pars$diff, pars$ma),
          seasonal=c(pars$sar, pars$sdiff, pars$sma))
  }, error = function(e) {
    # Return an error indicator, here we choose NULL for simplicity
    NULL
  })

  # Check if fit is a model object (success) or NULL (error)
  if (!is.null(fit)) {
    # Only calculate BIC for successful model fits
    nd_bic[i] <- BIC(fit)
  }  # If fit is NULL, the BIC remains Inf
}

# Find the parameters with the minimum BIC, excluding Inf values
best_model_index <- which.min(nd_bic)
nd_pars[best_model_index, ]


fit_pre <- Arima(pre_law, order = c(0, 1, 1), seasonal = c(0, 1, 1))
summary(fit_pre)
```
### Residual Analysis

The standardized residuals plot for the SARIMA model indicates the 
residuals are homoscedastic and center around zero mean. The ACF and PACF 
correlograms show that the residuals appear to look like white noise. All of the
p-values for the Ljung-Box statistic are above 0.05 at lags 1-20,suggesting the 
residuals do not show significant autocorrelation, and our model has adequately 
captured the autocorrelations and seasonality in the data. Lastly, the normal 
Q-Q plot only have small deviations from the 45 degrees line,overall, the 
$SARIMA (0,1,1)(0,1,1)_{12}$ model seem to have taken care of the significant 
spikes at lag 1-3 and the seasonality.

```{r, warning=FALSE, echo=FALSE}
res_pre <- scale(fit_pre$residuals)


# Check if residuals behave like white noise
plot(res_pre, xlab="Year", 
     main="Standarized ARIMA Residuals\n",
     ylab="",
     col="blue")
```

\newpage

```{r, warning=FALSE, echo=FALSE}
par(mfrow = c(2, 2))
# ACF and PACF
Acf(res_pre, main="ACF of SARIMA Residuals")
Pacf(res_pre, main="PACF of SARIMA Residuals")

# Q-Q plot
qqnorm(res_pre, main="Normal Q-Q Plot for Std Residuals")
qqline(res_pre)

# Ljung-Box test
p_values <- sapply(1:20, function(lag) {
  Box.test(res_pre, lag = lag, type = "Ljung-Box")$p.value
})

# Plot the p-values for each lag
plot(1:20, p_values, type="p", lwd = 2, col = "blue", 
     xlab = "Lag", ylab = "P-Value", main = "Ljung-Box Test P-Values (ARIMA)")
abline(h = 0.05, col = "red", lty = 2)  # Reference line for p-value = 0.05
```

## 2.5 Intervention Analysis
With an added intervention component representing the enactment of the seat belt
law in 1983, the model indicates a statistically significant impact of the law 
on reducing driver deaths. The intervention coefficient of -0.2278, with a very 
low p-value (approximately 0.000057), strongly suggests that the introduction of 
the seat belt law was associated with a decrease in fatalities. This result, 
given the robustness of the ARIMA modeling framework, underscores the 
effectiveness of the seat belt legislation in improving road safety.

```{r, warning=FALSE, echo=FALSE}
intervention <- ifelse(time(UKDriverDeaths) >= 1983, 1, 0)
intervention <- ts(intervention, start=c(1969, 1), frequency=12)
fit_post <- Arima(UKDriverDeaths,order=c(0,1,1), seasonal=c(0,1,1),
                               xreg=intervention)
summary(fit_post)

# Calculate the t-statistic
t_statistic <- -0.2278 / 0.0553

# Calculate the p-value for the two-tailed test
p_value <- 2 * pt(-abs(t_statistic), df=length(UKDriverDeaths)-
                    length(fit_post$coef))

# Print the p-value
p_value
```

### Residual Analysis
These plots show the similar results to fitting with pre-law data. The 
$SARIMA (0,1,1)(0,1,1)_{12}$ model with intervention seems to have taken care of 
the autocorrelation and seasonality.

```{r, warning=FALSE, echo=FALSE}
res_post <- scale(fit_post$residuals)

# Check if residuals behave like white noise
plot(res_post, xlab="Year", 
     main="Standarized ARIMA Residuals",
     ylab="",
     col="blue")
```

\newpage

```{r, warning=FALSE, echo=FALSE}
par(mfrow = c(2, 2))
# ACF and PACF
Acf(res_post, main="ACF of SARIMA Residuals")
Pacf(res_post, main="PACF of SARIMA Residuals")

# Q-Q plot
qqnorm(res_post, main="Normal Q-Q Plot for Std Residuals (SARIMA)")
qqline(res_post)


# Ljung-Box test
p_values <- sapply(1:20, function(lag) {
  Box.test(res_post, lag = lag, type = "Ljung-Box")$p.value
})

# Plot the p-values for each lag
plot(1:20, p_values, type="p", lwd = 2, col = "blue", 
     xlab = "Lag", ylab = "P-Value", main = "Ljung-Box Test P-Values (ARIMA)")
abline(h = 0.05, col = "red", lty = 2)  # Reference line for p-value = 0.05


```

## 2.6 Forcasting

The plot presents actual historical data in blue solid line, with fitted values 
from apost-law implementation model in red dashed line and forecasts from a 
pre-law model in green dotdash line. Accompanying this forecast, the orange 
dashed lines depict an 80% prediction interval, indicating the range within which
future values are expected to fall with 80% probability.

The close tracking of the fitted values to the actual data illustrates the 
post-law model's accuracy in capturing the impact of the seat belt legislation. 
In contrast, the forecast, extending beyond the observed data, projects what 
might have occurred without the law's influence, showing a clear deviation from 
the fitted values. This divergence, where the forecast predicts higher fatalities
than the observed and fitted data, suggests that the introduction of the seat 
belt law had a tangible, beneficial effect in reducing the number of driver 
deaths.

```{r, warning=FALSE, echo=FALSE}

# Plot the actual data
plot(exp(UKDriverDeaths), type="l", col="blue", xlab="Time", ylab="Deaths", main="UK Driver Deaths: Actual, Fitted, and Forecast")

# Add fitted values from the model with intervention
lines(exp(fitted(fit_post)), col="red", lty=2)

# Forecast the future values from the pre-intervention model
future_forecast <- forecast(fit_pre, h=24)

# Add the forecast from the pre-intervention model
lines(exp(future_forecast$mean), col="green", lty=4)

# Add 80% prediction limit for the forecast
lines(exp(ts(future_forecast$lower[1:24,1],start = c(1983, 1), frequency = 12)),
      col="orange", lty=5, lwd=2)

lines(exp(ts(future_forecast$upper[1:24,1],start = c(1983, 1), frequency = 12)),
      col="orange", lty=5, lwd=2)
# Add a legend to the plot
legend("topright", legend=c("Actual", "Fitted values by post-law model", 
                            "Forecast from pre-law model", 
                            '80% prediction limit for the forecast'), 
                            col=c("blue", "red", "green", 'orange'), 
                            lty=c(1,2,4,5),
                            lwd=c(1,1,1,3),
                            cex=0.8)

```

